Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-04-19 _manuscript/manuscript.qmd
BMS updates to GPT HPO manuscript

---
title: Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - HPO
  - chatGPT
  - rare disease 
bibliography: references.bib
---

```{r setup}
library(dplyr)
library(ggplot2)
library(patchwork)
library(HPOExplorer)
```

@@

The Human phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised database for phenotypic abnormalities. However, there is currently no information pertaining to the severity of each phenotype, making systematic analyses and prioritisation of results difficult. With 18,082 abnormalities now corresponding to over 10,000 rare diseases, manual curation of such phenotypic annotations by experts would be labor-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 model with Python to systematically annotate the severity of ~ 17,000 phenotypic abnormalities in the HPO. By checking that phenotypes with guaranteed outcomes were appropriately annotated, we demonstrate the potential for natural language processing technologies to automate the curation process accurately. For example, phenotypes such as "decreased male fertility" were used to compute a true positive rate, as they would be expected to be annotated as often, if not always, causing reduced fertility. Across the annotated outcomes, we observed > 73 % annotation accuracy. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence. These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human well being, and best targets for therapeutic intervention.

@@

Comprehensive annotation of phenotypic abnormalities is invaluable for defining, diagnosing, prognosing, and treating human disease. Since 2008, the Human phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes @Gargano2024-nj. Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now ~ 18,000 phenotypic abnormalities, corresponding to > 10,000 rare diseases, described. In recent years, the HPO has expanded its disease annotations so that each HPO term can have metadata including typical age of onset and frequency. In addition, there are the Clinical modifier (put this in italics) and Clinical course (also italics) subontologies, which contains terms to describe factors including severity and triggers, and mortality and progression, respectively. Describing the severity-related attributes of a disease is crucial for attaining significant objectives in rare diseases. This includes enhancing diagnostic capabilities, as well as prioritising and guiding gene therapy trials.

To date, the HPO has largely been manually curated by experts including clinicians, clinical geneticists, and researchers. Although this approach ensures the quality and accuracy of the ontology, it is time-consuming and labour-intensive. As a result, less than 1% of terms within the HPO have metadata pertaining to their features such as time course and severity. As artificial intelligence (AI) capabilities advance, there is an opportunity to integrate natural language processing technologies into assisting in the curation process. Here, we have used the OpenAI GPT-4 model (https://openai.com/) with Python to systematically annotate the severity of > 17,000 phenotypic abnormalities within the HPO. Our severity annotation framework was developed based on previously defined criteria developed through consultation with clinicians 2. The authors consulted 192 healthcare professionals for their opinions on the relative severity of various clinical characteristics: they used this to create a system for categorising the severity of  diseases. Briefly, each healthcare professional was sent a survey asking them to first rate how important a disease characteristic was for determining disease severity, and then to rate the severity of a set of given disease. Using the responses, the authors were able to categorise clinical characteristics into 4 ‘severity tiers’. While characteristics such as shortened lifespan in infancy and intellectual disability were identified as highly severe and placed into tier 1, sensory impairment and reduced lifespan were categorised as less severe and placed into tier 4. Being able to quickly ascribe severity measures based on those criteria to HPO phenotypes, will assist with interpreting phenome-wide studies.

Almost 800 phenotypes were annotated twice to evaluate annotation consistency, and a true positive rate of annotations was calculated to assess annotation accuracy. Additionally, based on the clinical characteristics and their occurrence, we have quantified the severity of each phenotype, providing an example of how these clinical characteristic annotations can be used to guide prioritisation of gene therapy trials. Ultimately, we hope that our resource will be of utility to those working in rare diseases, as well as the wider rare disease community.

@@

### Annotating the HPO using OpenAI GPT-4
We employed the OpenAI GPT-4 model with Python to annotate 16,982 terms within the Human phenotype Ontology (HPO). Our annotation framework was developed based on previously defined criteria for classifying disease severity @Lazarin2014-rz. We sought to evaluate the impact of phenotypes on outcomes including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset. Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response. For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred. Furthermore, our prompt design revealed that the optimal trade-off between the number of phenotypes and performance was achieved when inputting no more than two or three phenotypes per prompt. An example prompt can be seen in Figure 1. The phenotype outcome occurrence varied by category. While > 50% of annotated phenotypes never caused blindness, cancer, immunodeficiency, intellectual disability, reduced fertility, and sensory impairment, a similar proportion often or always had congenital onset (Fig. 1).

```{r occurence_plot}
#| echo: false
#| warning: false

# read in hpo gpt annotations 
hpo_annot <- HPOExplorer::gpt_annot_read()

# subset to clinical characteristics 
occurr_df <- hpo_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
                          "sensory_impairments", "immunodeficiency", "cancer", "reduced_fertility",
                          "congenital_onset")]

@@

# Loop through each column of interest and count the occurrences of 'always', 'often', 'rarely', 'never'
for (col in cols_of_interest) {
  counts <- table(occurr_df[[col]])
  count_df[col, ] <- counts[match(colnames(count_df), names(counts))]
}

@@

count_df_long <- count_df_long %>%
  dplyr::group_by(category) %>%
  dplyr::mutate(percentage = count / sum(count) * 100)

# set annotation order for plots 
annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
                      "congenital onset", "physical malformations", "sensory impairments", 
                      "immunodeficiency", "cancer", "reduced fertility")

count_df_long$category <- factor(count_df_long$category, levels = annotation_order)

# Stacked bar plot of clinical characteristic occurrence
ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(title = "Clinical characteristic occurrence",
       x = "", y = "HPO phenotypes (n)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "right", plot.title = element_text(hjust=0.5), 
        text = element_text(size=16),
        axis.text.x = element_text(angle=45, size=16, hjust=1), 
        title = element_text(size=15)) +
  scale_fill_brewer(palette = "GnBu", direction = -1)

```

![Figure 1: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics. a An example of the prompt input given to to GPT-4. The phenotypes listed in the last sentence were changed to allow ~17,000 phenotypes to be annotated. b Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all ~17,000 phenotypes annotated.The terms shown on the X-axis are the clinical characteristics, for which GPT-4 was asked to determine whether each phenotype caused them. For example, it annotated ‘Respiratory arrest’ as always causing death. Only 75 phenotypes were annotated as always causing death, despite 9,544 phenotypes often or rarely causing it.](figures/clinical_characteristic_occurrence.png)
### Annotation consistency and accuracy
To assess annotation consistency, 793 phenotypes underwent duplicate annotations, with two metrics employed to determine the 'consistency rate'. The first, less stringent metric, defined consistency as the duplicate annotations being either 'always' and 'often', or 'never' and 'rarely'. The second, more stringent metric, required exact agreement in annotation occurrences, e.g. 'always' and 'always'. For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least 80%, and for the more stringent metric, the lowest consistency rate was 57%, for congenital onset. An example of where annotations were inconsistent was for the HPO term ‘Acute leukaemia’. One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that ‘weakness and fatigue from leukaemia and its treatment can impair mobility’. The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that ‘acute leukaemia rarely impairs mobility directly’. Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, here is an example of where it failed to do so.

```{r annot_checks}
# Annotation checks
checks <- HPOExplorer::gpt_annot_check()

# Define items to be checked
items <- c("consistency_count", "consistency_rate",
           "consistency_stringent_count", "consistency_stringent_rate",
           "true_pos_count", "true_pos_rate")

metric_types <- c("Rate", "Count")[1]

@@

check_df$annotation <- gsub("_", " ", check_df$annotation)
annotation_order <- unique(check_df$annotation)
check_df$annotation <- factor(check_df$annotation, levels = annotation_order)
check_df$metric_category <- gsub("_", " ", check_df$metric_category)

metric_categories <- c("consistency", "consistency stringent")

```

```{r figure_2}
#| label: fig-checks
#| fig-cap: GPT-4 annotations are consistent and accurate across annotations. a Barplot showing the annotation consistency for 793 phenotypes that were annotated twice. On the left hand side, annotations were defined as consistent if the given phenotype was annotated to the same outcome occurrence both times, or if it was annotated to ‘always’ and ‘often’, or, ‘never’ and ‘rarely. On the right hand side, annotations were only defined as consistent if the given phenotype was annotated to the same outcome occurrence both times. b Bar plot of the true positive rate for each annotation. The true positives were determined based on the assumption that all descendants of some HPO phenotypes should be related to those outcomes. For instance, all descendents of the HPO term for “Neoplasm” were considered to be true positives for cancer (an assumption which is not always the case, as GPT-4 detects, as some neoplasms are benign). The labels above each bar indicate the number of phenotypes tested. 
#| fig-height: 4
#| fig-width: 6
consistency_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
                           aes(x = annotation, y = value,
                               fill = metric_category,
                               label = round(value, 2))) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  facet_grid(facets = ~metric_category, scales = "free") +
  theme_bw() +
  labs(x = "", y = "Value") +
  ggtitle("Annotation consistency") +
  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        legend.position = "none",
        text = element_text(size = 16),
        title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(x = 7.5, y = 1.05, label = "N phenotypes = 793"), size = 4)

metric_categories <- "true pos"

accuracy_plot <- ggplot(check_df[metric_type %in% metric_types & metric_category %in% metric_categories, ],
                        aes(x = annotation, y = value,
                            fill = metric_category,
                            label = n)) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  theme_bw() +
  labs(x = "", y = "Value") +
  ggtitle("True positive rate") +
  scale_fill_manual(values = c("#B1D5BBFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
        legend.position = "none",
        text = element_text(size = 16),
        title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5)) +
  geom_label(fill = "white", size = 4)

consistency_plot + accuracy_plot + plot_layout(widths = c(1, 0.7)) +
  plot_annotation(tag_levels = 'a') & theme(plot.tag = element_text(size = 18, face = "bold"))
```

In order to evaluate the accuracy of the annotations, we calculated a true positive rate. This involved identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the presence of certain conditions. For instance, the phenotypes 'decreased fertility in females' and 'decreased fertility in males' should, in the annotations, often, if not always, cause reduced fertility. We observed an encouraging true positive rate exceeding 70% across all phenotype outcomes. The lowest rate was observed for cancer (73%), for which all descendants of the HPO term ‘Neoplasm’ were considered to be true positives. However, this assumption, as GPT-4 detects, is not always true as these abnormal tissue/cell growths can be either cancerous or benign. For example, the HPO term ‘multiple digital exostoses’, which are benign bone tumours, is a descendant of the HPO term ‘Neoplasm’, and GPT-4 annotated it as rarely causing cancer, giving the justification that ‘Bone growths may rarely become cancerous’. This high level of accuracy underscores the robustness of our annotations and the reliability of the HPO framework in capturing clinically relevant phenotypic information. However, we acknowledge that the number of testable phenotypes for some of these categories are low. Furthermore, some of the ‘true positive’ phenotypes are synonymous to the clinical characteristic itself, for instance, for testing the true positive rate of the characteristic ‘Intellectual disability’, the phenotype ‘Severe intellectual disability’ was used.

>>>>

---
title: Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology
author:
  - name: Kitty B Murphy
    orcid: 0000-0002-8669-3076
    corresponding: true
    email: kitty.murphy@ukdri.ac.uk
    roles:
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Brian M Schilder
    orcid: 0000-0001-5949-2191
    corresponding: false
    email: brian_schilder@alumni.brown.edu
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
  - name: Nathan G Skene
    orcid: 0000-0002-6807-3180
    email: n.skene@imperial.ac.uk
    corresponding: true
    roles: 
      - Investigation
      - Project administration
      - Software
      - Visualization
    affiliations:
      - Department of Brain Sciences, Imperial College London, UK 
      - UK Dementia Research Institute at Imperial College London, UK
keywords:
  - Ontology
  - Human Phenotype Ontology
  - Large Language Model
  - GPT-4
  - Artificial intelligence
  - Generative AI
  - Rare disease 
  - Automation
  - Medicine
bibliography: references.bib
date: last-modified
citation:
  container-title: medRxiv
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
library(dplyr)
library(ggplot2)
library(ggtext)
library(patchwork)
library(HPOExplorer)
library(KGExplorer)
library(data.table)
```

```{r import-data}
#### Load HPO data
tag <- "v2024-02-08"
save_dir <- "data"
keep_descendants <- "Phenotypic abnormality"
p2g <- HPOExplorer::load_phenotype_to_genes(tag = tag,
                                            save_dir=save_dir)
hpo <- HPOExplorer::get_hpo(method="github",
                            tag=tag, 
                            save_dir=save_dir)
hpo <- KGExplorer::filter_ontology(ont = hpo, 
                                   keep_descendants = keep_descendants)
#### Read HPO GPT annotations 
gpt_annot <- HPOExplorer::gpt_annot_read(hpo = hpo)
#### Compute IDs counts
ids_matched <- intersect(c(hpo@terms,names(hpo@alternative_terms)), 
                         gpt_annot$hpo_id)
names_matched <- intersect(hpo@elementMetadata$name,
                           trimws(gsub("obsolete","",gpt_annot$hpo_name, ignore.case = TRUE)))

n_ids_hpo <- hpo@n_terms
n_ids_gpt <- length(unique(fcoalesce(gpt_annot$hpo_name,gpt_annot$hpo_id)))
```

```{r gpt_annot_check}
query_hits <- HPOExplorer::search_hpo(search_cols = c("name"),
                                      hpo = hpo)
#### Run annotation checks
checks <- HPOExplorer::gpt_annot_check(annot = gpt_annot, 
                                       query_hits = query_hits)
# checks$true_pos_rate
```

```{r gpt_annot_codify}
res_coded <- HPOExplorer::gpt_annot_codify(annot = gpt_annot)
annot_melt <- HPOExplorer::gpt_annot_melt(res_coded = res_coded)
```

```{r gpt_annot_plot}
gpt_annot_plot_out <- HPOExplorer::gpt_annot_plot(
  annot = gpt_annot[hpo_name!=keep_descendants],
  keep_descendants = keep_descendants, 
  width = 10)
```

@@

The Human Phenotype Ontology (HPO) has played a crucial role in defining, diagnosing, prognosing, and treating human diseases by providing a standardised framework for describing phenotypic abnormalities. However, documentation on the severity of each phenotype is currently very sparse, making systematic analyses and prioritisation of results challenging. With \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` rare diseases, manual curation of such phenotypic annotations by experts would be exceedingly labour-intensive and time-consuming. Leveraging advances in artificial intelligence, we employed the OpenAI GPT-4 large language model (LLM) to systematically annotate the severity of all phenotypic abnormalities in the HPO. First, we benchmarked the generative LLM annotations against ground-truth labels within the HPO (e.g. phenotypes in the 'Cancer' HPO branch were annotating as causing cancer by GPT-4). True positive recall rates across different annotations ranged from `r round(min(checks$true_pos_rate)*100)`-`r round(max(checks$true_pos_rate)*100)`% (mean=`r round(mean(checks$true_pos_rate)*100)`%), clearly demonstrating the ability of GPT-4 to automate the curation process with a high degree of fidelity. Using a novel approach, we developed a severity scoring system that incorporates both the nature of the phenotype outcome and the frequency of its occurrence. These severity metrics will enable efforts to systematically prioritise which human phenotypes are most detrimental to human health, and best targets for therapeutic intervention.

@@

Comprehensive annotation of phenotypic abnormalities is invaluable for defining, diagnosing, prognosing, and treating human disease. Since 2008, the Human Phenotype Ontology (HPO) has been instrumental to this, by providing a standardised database for the description and analysis of human phenotypes [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024]. Through developing open community resources, the depth and breadth of the HPO has continued to expand and there are now \>`r format(round(n_ids_hpo/100)*100, scientific = FALSE)` phenotypic abnormalities across \>`r format(round(length(unique(p2g$disease_id))/100)*100, scientific = FALSE)` diseases, described. In recent years, the HPO has expanded its disease annotations so that each HPO term can have metadata including typical age of onset and frequency. In addition, there are the *Clinical modifier* and *Clinical course* subontologies, which contains terms to describe factors including severity, triggers, mortality and progression, respectively. Describing the severity-related attributes of a disease is crucial for both research and clinical care of individuals with rare diseases. This includes enhancing diagnostic capabilities, as well as prioritising and guiding gene therapy trials.

To date, the HPO has largely relied on manual curation by experts including clinicians, clinical geneticists and researchers. Although this approach ensures the quality and accuracy of the ontology, it is time-consuming and labour-intensive. As a result, less than 1% of terms within the HPO have metadata pertaining to their features such as time course and severity. Artificial intelligence (AI) capabilities have advanced considerably in recent years, presenting new opportunities to integrate natural language processing technologies into assisting in the curation process. Specifically, there have recently been considerable advances in large language model (LLM) and their application to biomedical problems, in some cases performing as well or better than human clinicians on standardised medical exams and patient diagnosis tasks [@vanveenAdaptedLargeLanguage2024; @boltonBioMedLM7BParameter2024; @zhangBiomedGPTUnifiedGeneralist2023; @labrakBioMistralCollectionOpenSource2024; @guDomainSpecificLanguageModel2021; @singhalLargeLanguageModels2023; @shinBioMegatronLargerBiomedical2020; @chengExploringPotentialGPT42023; @singhalLargeLanguageModels2023; @oneilPhenomicsAssistantInterface2024; @luoBioGPTGenerativePretrained2022; @mcduffAccurateDifferentialDiagnosis2023; @singhalExpertLevelMedicalQuestion2023]. Recent work has demonstrated that the Generative Pre-trained Transformer 4 (GPT-4) foundation model [@openaiGPT4TechnicalReport2024], when combined with strategic prompt engineering, can outperform even specialist LLMs that are explicitly fine-tuned for biomedical tasks [@noriCanGeneralistFoundation2023]. In a landmark achievement, GPT-4 was the first LLM to surpass a score of 90% in the United States Medical Licensing Examination (USML) [@noriCanGeneralistFoundation2023].

Here, we have used GPT-4 to systematically annotate the severity of `r n_ids_gpt` / `r n_ids_hpo` (`r round(n_ids_gpt/n_ids_hpo*100,1)`%) phenotypic abnormalities within the HPO. Our severity annotation framework was adapted from previously defined criteria developed through consultation with clinicians [@Lazarin2014-rz]. The authors consulted 192 healthcare professionals for their opinions on the relative severity of various clinical characteristics: they used this to create a system for categorising the severity of diseases. Briefly, each healthcare professional was sent a survey asking them to first rate how important a disease characteristic was for determining disease severity, and then to rate the severity of a set of given disease. Using the responses, the authors were able to categorise clinical characteristics into 4 'severity tiers'. While characteristics such as shortened lifespan in infancy and intellectual disability were identified as highly severe and placed into tier 1, sensory impairment and reduced lifespan were categorised as less severe and placed into tier 4. Standardised metrics of severity allow clinicians to quickly assess the urgency of treating a given phenotype, as well as prognosing what outcomes might be expected.

`r checks$consistency_count[[1]]` phenotypes were annotated multiple times to evaluate the consistency of responses generated by GPT-4. For a subset of phenotypes with known expected annotations, true positive rates were calculated to assess recall. Additionally, based on the clinical characteristics and their occurrence, we have quantified the severity of each phenotype, providing an example of how these clinical characteristic annotations can be used to guide prioritisation of gene therapy trials. Ultimately, we hope that our resource will be of utility to those working in rare diseases, as well as the wider healthcare community.

@@

### Annotating the HPO using GPT-4

```{r fig-occurrence}
#| label: fig-occurrence
#| fig-cap: GPT-4 was able to annotate all human phenotypes based on whether they are always/often/rarely/never associated with different clinical characteristics. **a**, An example of the prompt input given to to GPT-4. The phenotypes listed in the second to last sentence (*italicised*) were changed to allow all HPO phenotypes to be annotated. **b**, Stacked bar plot showing the proportion of the occurrence of each clinical characteristic across all annotated HPO phenotypes. The terms shown on the x-axis are the clinical characteristics for which GPT-4 was asked to determine whether each phenotype caused them. 
#| fig-height: 7
#| fig-width: 8

# subset to clinical characteristics 
occurr_df <- gpt_annot[,c("hpo_name", "intellectual_disability", "death", "impaired_mobility", "blindness",
                          "sensory_impairments", "immunodeficiency", "cancer", "reduced_fertility",
                          "congenital_onset")]

@@

## Loop through each column of interest and count the occurrences of:
## 'always', 'often', 'rarely', 'never'
for (col in cols_of_interest) {
  counts <- table(occurr_df[[col]])
  count_df[col, ] <- counts[match(colnames(count_df), names(counts))]
}

@@

count_df_long <- count_df_long |>
  dplyr::group_by(category) |>
  dplyr::mutate(percentage = count / sum(count) * 100)

# set annotation order for plots 
annotation_order <- c("death", "intellectual disability", "impaired mobility", "blindness", 
                      "congenital onset", "physical malformations", "sensory impairments", 
                      "immunodeficiency", "cancer", "reduced fertility")
count_df_long$category <- factor(count_df_long$category, 
                                 levels = annotation_order)

# Load the prompt generation function 
source("code/prompt_gen.R")
prompts <- prompt_gen_table(hpo = hpo)
txt_prompt <- prompts$prompt[1688]
for(x in prompts$terms[[1688]]){
  txt_prompt <- sub(x,paste0("*",x,"*"),txt_prompt)
}
for(x in c(
  eval(formals(prompt_gen_table)$effects),
  eval(formals(prompt_gen_table)$responses),
  "justification")){
  txt_prompt <- sub(x,paste0("**",x,"**"),txt_prompt)
}
## Subplot a: Example prompt
f1a <- ggplot() + 
  labs(title="Example GPT-4 prompt") +
  ggtext::geom_textbox(aes(x=.5,y=1),
                       box.padding = margin(20, 10, 20, 10),
                       box.margin = margin(0, 0, 0, 0),
                       box.r = grid::unit(30, "pt"),
                       width = unit(.8, "npc"),
                       halign=.5,
                       fill="grey90",
                       label=txt_prompt) +
  theme_void() +
  theme(plot.title = element_text(hjust=0.5))

## Subplot b: Stacked bar plot of clinical characteristic occurrence.
f1b <- ggplot(count_df_long, aes(x = category, y = count, fill = condition)) +
  geom_bar(stat = "identity") +
  labs(title = "Clinical characteristic occurrence",
       x = "", y = "HPO phenotypes (n)", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "right", 
        plot.title = element_text(hjust=0.5),  
        axis.text.x = element_text(angle=45, hjust=1),  
        ) +
  scale_fill_brewer(palette = "GnBu", direction = -1)

(
  f1a| 
    (f1b/plot_spacer() ) + plot_layout(heights = c(1, .1))
) + 
  patchwork::plot_layout(widths = c(2,1)) +
  patchwork::plot_annotation(tag_levels = letters) 
```

We employed the OpenAI GPT-4 model with Python to annotate `r length(unique(gpt_annot$hpo_name))` terms within the HPO (`r tag`) [@kohlerHumanPhenotypeOntology2021; @garganoHumanPhenotypeOntology2024]. Our annotation framework was developed based on previously defined criteria for classifying disease severity [@Lazarin2014-rz]. We sought to evaluate the impact of phenotypes on factors including intellectual disability, death, impaired mobility, physical malformations, blindness, sensory impairments, immunodeficiency, cancer, reduced fertility, and congenital onset. Through prompt design we found that the performance of GPT-4 improved when we incorporated a scale associated with each effect and required a justification for each response. For each effect, we asked about the likelihood of its occurrence - whether it never, rarely, often, or always occurred.

```{r always_death}
never_thresh <- 50
never_dt <- subset(count_df_long, condition=="never" & percentage>=never_thresh)
always_death <- gpt_annot[,list(always_death=all(death=="always"),
                                hpo_name=hpo_name[1]),
                          by="hpo_id"][always_death==TRUE,]
```

Phenotype outcome occurrence varied across annotation categories. \>`r never_thresh`% of phenotypes never caused `r paste(never_dt$category[-1], collapse=", ")` or `r never_dt$category[1]`. Only a minority of phenotypes (`r round(sum(subset(count_df_long, condition %in% c("never") & category=="congenital onset")$percentage),1)`%) never had a congenital onset, which is expected as rare disorders tend to be early onset genetic conditions [Fig. @fig-occurrence].

Less than `r ceiling(subset(count_df_long, condition=="always" & category=="death")$percentage)`% of phenotype always directly resulted in death, such as 'Stillbirth', 'Anencephaly' and 'Bilateral lung agenesis'. Only `r uniqueN(gpt_annot[death=="always"]$hpo_name)` phenotypes were annotated as always causing death, while `r uniqueN(gpt_annot[death %in% c("often","rarely")]$hpo_name)` phenotypes were annotated as often or rarely causing death. `r uniqueN(gpt_annot[death %in% c("never")]$hpo_name)` phenotypes were annotated as never causing death. This included `r sum(grepl("syndactyly",gpt_annot[death %in% c("never")]$hpo_name, ignore.case = TRUE))` unique forms of syndactyly, a non-lethal condition that causes fused or webbed fingers (occurring 1 in 1,200–15,000 live births). While not life-threatening itself, syndactyly is a symptom of genetic disorders that can cause life-threatening cardiovascular and neurodevelopmental defects, such as Apert Syndrome [@garagnaniSyndromesAssociatedSyndactyly2013]. This example highlights the ability of GPT-4 to successfully distinguish between phenotypes that directly cause lethality, and those that are often associated with diseases that cause lethality.

### Annotation consistency and recall

To assess annotation consistency, `r checks$consistency_count[[1]]` phenotypes underwent duplicate annotations. We employed two different metrics to determine the *consistency rate*. The first, less stringent metric, defined consistency as the duplicate annotations being either 'always' and 'often', or 'never' and 'rarely'. The second, more stringent metric, required exact agreement in annotation occurrences, e.g. 'always' and 'always'. For the less stringent metric, duplicated phenotypes were annotated consistently at a rate of at least `r floor(min(checks$consistency_rate)*100)`%, and for the more stringent metric, the lowest consistency rate was `r round(min(checks$consistency_stringent_rate)*100)`% for congenital onset. An example of where annotations were inconsistent was for the HPO term 'Acute leukaemia'. One time, GPT-4 annotated it as often causing impaired mobility, giving the justification that 'weakness and fatigue from leukaemia and its treatment can impair mobility'. The other time, GPT-4 annotated it as rarely causing impaired mobility, giving the justification that 'acute leukaemia rarely impairs mobility directly'. Despite specifying in the prompt for GPT-4 not to take into consideration indirect effects, here is an example of where it failed to do so.

```{r annot_checks}
# Define items to be checked
items <- c("consistency_count", "consistency_rate",
           "consistency_stringent_count", "consistency_stringent_rate",
           "true_pos_count", "true_pos_rate")
metric_types <- c("Rate", "Count")[1]

@@

check_df$annotation <- gsub("_", " ", check_df$annotation)
annotation_order <- unique(check_df$annotation)
check_df$annotation <- factor(check_df$annotation, levels = annotation_order)
check_df$metric_category <- gsub("_", " ", check_df$metric_category)
check_df[metric_category=="consistency",metric_category:="consistency lenient"]
```

```{r fig-checks}
#| label: fig-checks
#| fig-cap: GPT-4 annotations are consistent and accurate across annotations. **a**, Barplot showing the annotation consistency within phenotypes that were annotated more than once. In the lenient metric, annotations were collapsed into two groups ('always'/'often' and 'never'/'rarely'). For a given clinical characteristic within a given phenotype, if an annotation was always within the same group it was considered consistent. In the stringent metric, all four annotation categories were considered to be different from one another. Thus, annotations were only defined as consistent if they were all identical. The blue dashed line indicates the probability of two annotations being consistent by chance in the lenient metric (~1/2). The gold dashed line is the probability of two annotations being consistent by chance in the stringent metric (~1/4). **b**, Bar plot of the true positive rate for each annotation. The labels above each bar indicate the number of phenotypes tested. 
#| fig-height: 5
#| fig-width: 6

lab_size <- 3
consistency_plot <- ggplot(check_df[metric_type %in% metric_types &
                                    grepl("consistency",metric_category)],
                           aes(x = annotation, y = value,
                               fill = stringr::str_to_sentence(
                                 gsub("consistency ", "", metric_category)
                                 ),
                               label = round(value, 2))) +
  geom_bar(stat = "identity",position = "dodge") +
  labs(title="Annotation consistency",
       x = NULL, y = "Mean consistency", 
       fill=NULL) +
  scale_fill_manual(values = c("#C3DAEAFF", "#ECE28BFF")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top", 
        strip.background = element_rect(fill="white"),
        plot.title = element_text(hjust = 0.5)) +
  geom_text(aes(x = 1, y = 1, label = paste0("n=",n," phenotypes")),
            size=lab_size,
            check_overlap=TRUE, hjust=0) +
  ## Compute the empirical probability of annotations being consistent by chance 
  ## with 2 options and N duplicates
  geom_hline(yintercept = mean(1/2^checks$annot$pheno_count), 
             color="dodgerblue", linetype="dashed")+
  ## Compute the empirical probability of annotations being consistent by chance 
  ## with 4 options and N duplicates
  geom_hline(yintercept = mean(1/4^checks$annot$pheno_count), 
             color="gold3", linetype="dashed")

recall_plot <- ggplot(check_df[metric_type %in% metric_types &
                               metric_category %in% c("true pos"), ],
                        aes(x = annotation, y = value,
                            fill = metric_category,
                            label = n#paste0("n\n=\n",n)
                        )
                      ) +
  geom_bar(stat = "identity", show.legend = TRUE) +
  theme_bw() +
  labs(x = NULL, y = "Recall") +
  ggtitle("True positive rate") +
  scale_fill_manual(values = c("#B1D5BBFF")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none", 
        plot.title = element_text(hjust = 0.5)) +
  geom_label(fill = alpha("white",1),
            size = lab_size)

consistency_plot + recall_plot + 
  plot_layout(widths = c(1, 1)) +
  plot_annotation(tag_levels = 'a')
```

In order to evaluate the validity of the annotations, we calculated a true positive rate. This involved identifying specific branches within the HPO that would contain phenotypes that would reliably indicate the presence of certain conditions. For instance, the phenotypes 'Decreased fertility in females' and 'Decreased fertility in males' should often or always cause reduced fertility. We observed an encouraging true positive rate exceeding `r floor(min(checks$true_pos_rate)*100)`% across in every clinical characteristic and achieving perfect recall (100%) in `r sum(checks$true_pos_rate==1)`/`r length(checks$true_pos_rate)` characteristics.

```{r}
lowest_recall <- names(checks$true_pos_rate)[checks$true_pos_rate==min(checks$true_pos_rate)]
lowest_recall_dt <- checks$annot[get(lowest_recall)=="never" & hpo_id %in% query_hits[[lowest_recall]]]
```

The lowest true positive rate was observed for `r gsub("_"," ",lowest_recall)`, with `r round(checks$true_pos_rate[[lowest_recall]]*100,1)`% recall across `r checks$true_pos_count[[lowest_recall]]` HPO phenotypes. Some cases in which the GPT-4 annotations disagreed with the HPO ground truth included: `r paste(shQuote(head(lowest_recall_dt$hpo_name, 6)), collapse=", ")`. In the case of `r shQuote(lowest_recall_dt$hpo_name[1])` it provided the justification that `r shQuote(lowest_recall_dt$physical_malformations_justification[1])`. In another instance, GPT-4 noted that `r shQuote(lowest_recall_dt$hpo_name[2])` is `r shQuote(lowest_recall_dt$physical_malformations_justification[2])`. This indicates that while technically incorrect according to our predefined benchmarks, a case could in fact be made that mild skin conditions do not rise to the level of physical malformations.
